---
title: "ggplot2 basics"
author: "John D. Lee"
date: "2/2/2020"
output:
  html_notebook
---

This code provides a high level overview of ggplot, with examples of:

* mapping and setting aesthetic properties

* local and global mapping of aesthetic properties and overlaying summary data

* building a plot by adding layers including, sina plot, statistical models and faceting

* ordering facets to make differences clearer



```{r, load_packages, echo=FALSE}
library(tidyverse) # For ggplot, dplyr, tidyr...
library(skimr) # For quick review of data
library(ggforce) # For sina plots
library(lme4) # For the sleep study data

set.seed(888)

```


## skimr, head, and tail to review the data

```{r review_data}

sleep.df = sleepstudy # Read data from package

skim(sleep.df) # Summarizes dataframe

head(sleep.df) # First rows of dataframe
tail(sleep.df) # Last rows of dataframe

```

## Simple scatterplot: Map variables to aesthetic properties of geometric elements
```{r simple_scatter}

ggplot(data = sleep.df, mapping = aes(x = Days, y = Reaction)) + 
  geom_point()

```


## Simple scatterplot: Setting aesthetic properties
A common error is to map rather than set a color to a geom.  These graphs show what happens.
```{r set_properties}

ggplot(data = sleep.df, mapping = aes(x = Days, y = Reaction)) + 
  geom_point(colour = "darkblue") 

ggplot(data = sleep.df, mapping = aes(x = Days, y = Reaction, colour = "darkblue")) + 
  geom_point() 

```


## Simple scatterplot: Layering geometric elements

Note that each layer is plotted on top of the previous layers. Here the violin plot lies beneath the sina plot.

```{r sina_layer}

ggplot(data = sleep.df, mapping = aes(x = as.factor(Days), y = Reaction)) + 
  geom_violin(fill= "grey75", colour = "grey75") +
  geom_sina(colour = "darkblue", size = .7)

```


## Simple scatterplot: Local and global mapping

Use global mapping to minimize typing, ease revisions, and minimize confusing ambiguity.

Local mapping also makes it possible to add points from another data set.

```{r mapping_local_global}

## Global
ggplot(data = sleep.df, mapping = aes(x = as.factor(Days), y = Reaction)) + 
  geom_violin(fill= "grey85", colour = "grey85") +
  geom_sina(colour = "darkblue", size = .7)

## Local
ggplot(data = sleep.df) + 
  geom_violin(mapping = aes(x = as.factor(Days), y = Reaction), fill= "grey85", colour = "grey85") +
  geom_sina(mapping = aes(x = as.factor(Days), y = Reaction), colour = "darkblue", size = .7)

## Local mapping to overlay data
m.sleep.df = sleep.df %>% 
  group_by(Days) %>% 
  summarise(m.reaction = mean(Reaction))

ggplot(data = sleep.df, mapping = aes(x = as.factor(Days), y = Reaction)) + 
  geom_violin(fill = "grey85", colour = "grey85") +
  geom_sina(colour = "darkblue", size = .7) +
  geom_point(data = m.sleep.df, aes(y = m.reaction), size = 3)

```


## Grouped data with linear regression: Slopes and intercepts both vary across people

Simply adding a layer calculates a regression line. When the data area grouped a line is calculated for each group.

Note the use of "group" to identify data that should be plotted as a line.

A common error is failing to identify the groups that define the lines.
```{r scatterplot_lm}

ggplot(sleep.df, aes(Days, Reaction, group = Subject)) +
  geom_line(alpha = .33) 

# What happens when you don't identify the groups?
ggplot(sleep.df, aes(Days, Reaction)) +
  geom_line(alpha = .33) 


ggplot(sleep.df, aes(x = Days, y = Reaction, group = Subject)) + 
  geom_line(alpha = .33) +
  geom_smooth(method = "lm", se = FALSE, size = .8)

# What happens when you don't identify the groups?
ggplot(sleep.df, aes(Days, Reaction)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```


## Faceted scatterplot: Small multiples to show individual responses

facet_grid adds a different sort of layer that creates a graph for each combination of data defined by the grid variables, in this case "Subject"
```{r facet_scatterplot_loess, fig.height=3.25, fig.width=10.25}

ggplot(data = sleep.df, mapping = aes(x = Days, y = Reaction)) + 
  geom_point() +
  facet_grid(.~Subject)

```


## Faceted Scatterplot with Loess Fit: A layer of loess fits shows a linear model doesn't fit all participants{.smaller}
```{r Scatterplot_loes, fig.height=3.25, fig.width=10.25}

ggplot(sleep.df, aes(Days, Reaction)) + 
  geom_point() +
  geom_smooth(method = "loess") +
  facet_grid(.~Subject)
```


## Faceted scatterplot with linear regression: A linear model shows slopes and intercepts differ{.smaller}
```{r facet_scatterplot_lm, fig.height=3.25, fig.width=10.25}

ggplot(sleep.df, aes(Days, Reaction)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(.~Subject)
```


## Faceted scatterplot with linear regression: Subjects ordered by mean reaction time
```{r facet_scatterplot_lm_reordered, warnings=FALSE, fig.height=3.25, fig.width=10.25}

ggplot(sleep.df, aes(Days, Reaction)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(.~reorder(x = Subject, X = Reaction, FUN = mean)) +
  scale_x_continuous(breaks = c(2, 4, 6, 8)) +
  labs(title = "Sleep deprevation slows reaction time",
       subtitle = "Individual participants ordered by mean reaction time",
       y = "Reaction time (s)", x = "Days into experiment")
```
