---
title: "Exploratory data analysis with ggplot and dplyr"
date: "2/7/2020"
output:
  html_notebook
---

This demo provides an overview of ggplot for exploratory data analysis, including:

* reading csv files

* assertions to check data

* select and transform data

* single dimensional distributions (e.g., histograms)

* two-dimensional distributions (e.g., scatterplots) 

* scatterplot matrices

* parallel  coordinate plot

The data used for these examples comes from a clean dataset that describes red wines in terms of ratings and physical properties:
(https://www.kaggle.com/uciml/red-wine-quality-cortez-et-al-2009)
(https://archive.ics.uci.edu/ml/datasets/wine+quality)


Bonus: reading multiple files

```{r load_packages, echo=FALSE}
library(tidyverse) # For ggplot, dplyr, tidyr...
library(skimr) # For quick review of data
library(assertr) # For checking data complies with expecations
library(ggforce) # For SPLOM and sina plots


set.seed(888) # To ensure consistent results from non-deterministic procedures
rm(list = ls()) # Removes all variables

```


## Read and skim data
One of the more tricky things can be identifying where the file is that you are reading.  It is best to create a directory that contains only your data and a directory that contains your notebooks. When you do that you will need to specify the path to your data.  The command below uses the ".." to indicate that the data are outside the directly where the notebook is. This can be confusing and the following post helps explain: http://jenrichmond.rbind.io/post/how-to-use-the-here-package/

The following works for a directory structure that looks like this:
DemosDataScienceforPeople
  /notebooks
  /data



```{r review_data}
wine.df = read_csv("../data/winequality-red.csv")
colnames(wine.df) <- make.names(colnames(wine.df), unique=TRUE) # Ensures names are valid and don't contain spaces

getwd()

skim(wine.df)

head(wine.df)

tail(wine.df)

```

## Assertions to check data

For more information see: https://docs.ropensci.org/assertr/

Assertions allow you to check whether the data comply with what you expect. Expectations can include falling with range of permissible values, absence of missing values, absence of extreme outliers.

One of the most useful features of the assertr package is the abiltiy to identify outliers using the Mahalanobis distance. This distance measure indicates rows that have a combination of points that puts it more than N median absolute deviations from the other points. Median absolute deviaions are a robust version of standard deviations and so large values indicate rows that are very different than the others.

The "everything()" selection method is one of many methods from dplyr for selecting variables. For a summary of other methods search RStudio help for select_helpers. For example, "contains("acid") would select all variables with "acid" in the name. You can also use something like this:"fixed.acidity:chlorides, density".This selects columns between "fixed.acidity" and "chlorides", as well as "density"

```{r assertions}

error.df = wine.df %>%
  chain_start %>%
  assert(in_set(1, 2, 3, 4, 5, 6, 7, 8), quality) %>% # Wine quality must be between 1 and 8
  assert_rows(num_row_NAs, within_bounds(0, 1), everything()) %>% # At most one NA in a row
  verify(fixed.acidity > 0 & fixed.acidity < 14) %>% # Acidity is between 1 and 10
  insist_rows(maha_dist, within_n_mads(8), everything()) %>% # Number of median absolute deviations
  chain_end(error_fun = error_df_return) 
error.df


wine_checked.df = wine.df %>% rowid_to_column("index")

wine_checked.df = left_join(wine_checked.df, error.df, by = "index") %>% 
  mutate(error = !is.na(verb))

ggplot(wine_checked.df, aes(quality, chlorides, colour = error)) +
  geom_point() +
  scale_color_manual(values = c("grey50", "red"))

```


# Select and transform variables

```{r select_transform}

wine.df = wine.df %>% select(-chlorides) # Selects a subset of variables (columns)

wine.df = wine.df %>% filter(fixed.acidity<12) # Filters a subset of observations (rows) 

```


##  Histograms

Be sure to vary the binwidth or the number of bins. Data you might think are continuous might be somewhat discrete.

```{r histogram}

ggplot(wine.df, aes(quality)) +
  geom_histogram()

ggplot(data = wine.df, mapping = aes(x = volatile.acidity )) +
  geom_histogram()

ggplot(data = wine.df, mapping = aes(x = volatile.acidity )) +
  geom_histogram(bins = 120)

ggplot(wine.df, aes(volatile.acidity )) +
  geom_histogram(bins = 120)

```

## Scatterplot

```{r scatterplot}

ggplot(data = wine.df, mapping = aes(x = volatile.acidity, y = quality)) +
  geom_point()

ggplot(data = wine.df, mapping = aes(x = volatile.acidity, y = quality)) +
  geom_point() +
  geom_smooth() 

ggplot(data = wine.df, mapping = aes(x = volatile.acidity, y = quality)) +
  geom_point() +
  geom_smooth(colour = "grey70") + 
  geom_smooth(method = "lm", se = FALSE) 

```


## Scatterplot matrix
https://www.data-imaginist.com/2019/a-flurry-of-facets/

```{r scatterplotmatrix}

ggplot(wine.df, aes(x = .panel_x, y = .panel_y)) +
  geom_point(size = .1) + #1
  geom_autodensity() + #2
  geom_density2d(colour = "grey70") + #3
  facet_matrix(vars(fixed.acidity:citric.acid), 
               layer.diag = 2, layer.upper = 3, 
               grid.y.diag = FALSE) +
  theme_minimal()


ggplot(wine_checked.df, aes(x = .panel_x, y = .panel_y)) +
  geom_point(aes(color = error), size = .1) + #1
  geom_autodensity() + #2
  geom_density2d(colour = "grey70") + #3
  facet_matrix(vars(fixed.acidity:citric.acid), 
               layer.diag = 2, layer.upper = 3, 
               grid.y.diag = FALSE) +
  scale_color_manual(values = c("grey50", "red")) +
  theme_minimal()


ggplot(wine.df, aes(x = .panel_x, y = .panel_y)) +
  geom_point(size = .1) + #1
  geom_autodensity() + #2
  geom_density2d(colour = "grey70") + #3
  facet_matrix(vars(fixed.acidity:quality), 
               layer.diag = 2, layer.upper = 3, 
               grid.y.diag = FALSE) +
  theme_minimal()


ggplot(wine_checked.df, aes(x = .panel_x, y = .panel_y)) +
  geom_point(aes(color = error), size = .1) + #1
  geom_autodensity() + #2
  geom_density2d(colour = "grey70") + #3
  facet_matrix(vars(fixed.acidity:quality), 
               layer.diag = 2, layer.upper = 3, 
               grid.y.diag = FALSE) +
  scale_color_manual(values = c("grey50", "red")) +
  theme_minimal()


```

## Parallel coordinate plot

A parallel coordinate plot highlights the distribution of the variables and trends across variables.  
```{r parallel_coordplot}

## Add indicator of high-quality wines
wine.df = wine.df %>%  
  mutate(hi_quality = if_else(quality> 6, "hi", "lo"))

## Create "long" version of data with column of variable names and one for variable values
long.wine.df = wine.df %>% mutate(id = row_number()) %>% 
  pivot_longer(fixed.acidity:quality, names_to = "variables", values_to = "values")


## Scaled values to z-scores
scaled.long.wine.df = long.wine.df %>% group_by(variables) %>% 
  mutate(s.values = scale(values))

## Parallel coordinate plots can be messy
ggplot(scaled.long.wine.df, aes(variables, s.values)) +
  geom_line(aes(group = id), size = .1, alpha = .2) +
  coord_flip()

## Highlighting cases of interest can reveal patterns
ggplot(scaled.long.wine.df, aes(variables, s.values)) +
  geom_line(aes(group = id, colour = hi_quality, size = hi_quality), alpha = .2) +
  scale_color_manual(values = c("red", "grey50")) +
  scale_size_manual(values = c(.3, .1)) +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "none")
```

## Multiple histograms and sina plots

```{r multidistribution}

ggplot(long.wine.df, aes(values)) +
  geom_histogram(bins = 75) +
  facet_wrap(~variables, nrow = 8, scales = "free")


ggplot(scaled.long.wine.df, aes(variables, s.values)) +
  geom_sina(size = .01, alpha = .5) +
  coord_flip()

```


## Save the data
It is important to keep the original raw data unchanged and to save any modified data to a new file.

```{r savedata}

wine.df %>% write_csv(path = "../data/refined_wine.csv")

```


# Bonus

Not needed for the assignment, but might be useful in the future

## Reading multiple files
Data files containing baby names can be downloaded from:
https://catalog.data.gov/dataset?tags=baby-names

Reading data from many files is a common challenge. How to read multiple files efficiently is demonstrate with the US baby names data. Data for most popular baby names is contained in csv files, one per year for the last 140 years.

```{r read_multiplefiles}

## Identify files containing data
data_directory = "../data/names"

data_files = list.files(data_directory, full.names = TRUE)

names(data_files) = list.files(data_directory) # Adds names to character vector

## Use map command to apply function "read_csv" to list of files
names.df = map_df(data_files, read_csv,
                   .id = "yob", quote = "",
                   col_names = c("name", "gender", "babies"),
                   col_types = list(col_character(), 
                                    col_factor(levels = c("M", "F")),
                                    col_integer()))

## Clean the date variable 
names.df = names.df %>% mutate(yob = str_replace(yob, "yob", "")) %>% 
  mutate(yob = str_replace(yob, ".txt", "")) %>% 
  mutate(yob = as.numeric(yob)) 
  

## Calculate percent of babies for each name
names.df = names.df %>% 
  group_by(yob) %>% 
  mutate(total_babies = sum(babies)) %>% 
  ungroup() %>% 
  mutate(percent = babies/total_babies)


## Plot trend over time
ggplot(names.df, 
       aes(yob, percent, group = interaction(name, gender), colour = name=="John")) +
  geom_line(size = .3, alpha = .4) +
  scale_color_manual(values = c("grey50", "red")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Popular names have become less dominant",
       caption = "data source: https://catalog.data.gov/dataset?tags=baby-names",
       x = "Year of birth",
       y = "Percent of babies with each name") +
  theme_bw()

```



